Topics in Statistical Theory --- Example Sheet 1

### 2020-11-18 distribution_theory/distribution_ES2.tex
aaarhhh
hellow

% Frequently used letters in the blackboard alphabet
\newcommand{\CC}{\Bb C}
\newcommand{\NN}{\Bb N}
\newcommand\ZZ{\Bb Z}
\newcommand{\PP}{\Bb P}
\newcommand{\QQ}{\Bb Q}
\newcommand{\RR}{\Bb R}
\newcommand{\KK}{\Bb K}
\newcommand{\DD}{\Cal D}
\newcommand\EE{\Cal E}
\renewcommand\SS{\Cal S}
\newcommand\FF{\Cal F}
\newcommand\vd{\vdot}
%\newcommand\ZZ{\Bb Z}

@@

\begin{document}
\maketitle
\begin{question}
Let $u, v \in \DD'(\RR^n)$, one of which has compact support. Show that the convolution $u * v$, defined as in your notes, is uniquely defined and gives rise to an element of $\DD'(\RR^n)$. 
\end{question}

\begin{proof}
	The convolution between $u, v \in \DD'(\RR^n)$ is defined by the formula
	\begin{equation} \label{def:convolution}
	(u * v) * \phi = u * (v * \phi) \quad\text{for all $\phi \in \DD(\RR^n)$}. 
	\end{equation}
	To show that this is uniquely defined, recall that for all $u \in \DD'(\RR^n) \phi\in \DD(\RR^n)$, we have $\ang{u, \phi} = (u * \refl\phi)(0)$. Therefore, we have
	\[
	\ang{u * v, \phi} = ((u * v) * \refl\phi)(0) = (u * (v * \refl\phi))(0),
	\]
	which shows that the formula \cref{def:convolution} uniquely defines $\ang{u * v, \phi}$ for any $\phi \in \DD(\RR^n)$, and therefore that $u * v$ is well-defined.

@@

	\TODO Give proof.

	Now, for $u \in H^s(\RR^n)$, let $(\phi_m) \to u$ in $L^2$. Then we have
	\begin{align*}
		\norm{\phi_m - u}_{H^s}^2 = \int \ang{\lambda}^{2s} \abs{(\phi_m - u)(\lambda)}^2 \dd{\lambda}
	\end{align*}
\end{proof}
%-----------------------------------------------------------------------------
\setcounter{question}{18}

>>>>

% Frequently used letters in the blackboard alphabet
\newcommand{\CC}{\Bb C}
\newcommand{\NN}{\Bb N}
\newcommand\ZZ{\Bb Z}
\newcommand{\PP}{\Bb P}
\newcommand{\QQ}{\Bb Q}
\newcommand{\RR}{\Bb R}
\newcommand{\KK}{\Bb K}
\newcommand{\DD}{\Cal D}
\newcommand\EE{\Cal E}
\renewcommand\SS{\Cal S}
\newcommand\FF{\Cal F}
\newcommand\vd{\vdot}
\newcommand\Rr{\Cal R}
%\newcommand\ZZ{\Bb Z}

@@

\begin{document}
\maketitle

\begin{mdframed}
	We will write $\Rr$ and $\FF$ for the reflection and Fourier transform operators. 
\end{mdframed}
\begin{question}
Let $u, v \in \DD'(\RR^n)$, one of which has compact support. Show that the convolution $u * v$, defined as in your notes, is uniquely defined and gives rise to an element of $\DD'(\RR^n)$. 
\end{question}

\begin{proof}
	The convolution between $u, v \in \DD'(\RR^n)$ is defined by the formula
	\begin{equation} \label{def:convolution}
	(u * v) * \phi = u * (v * \phi) \quad\text{for all $\phi \in \DD(\RR^n)$}. 
	\end{equation}
Recall that for all $u \in \DD'(\RR^n) \phi\in \DD(\RR^n)$, we have $\ang{u, \phi} = (u * \refl\phi)(0)$, and therefore $u * v$ should satisfy
\[
\ang{u * v, \phi} = ((u * v) * \refl\phi)(0) = (u * (v * \refl\phi))(0) = \ang{u, \Rr(v * \refl\phi)}. 
\] Therefore 
\begin{enumerate}
	\item Define $w \colon \DD(\RR) \to \CC$ by 
	\[
	\ang{w, \phi} = \ang{u, \Rr(v * \refl\phi)}, 
	\]
	then we will show that $w$ satisfies $w * \phi = u * (v * \phi)$ for all $\phi \in \DD(\RR)$. 
	\begin{align*}
		(w * \phi)(x) &= \ang{w, \tau_x\refl\phi} = \ang{u, \Rr(v * \Rr(\tau_x \refl\phi))} = \ang{u, \Rr(v * \tau_{-x}\phi)} \\
		&= \ang{u, \Rr \tau_{-x} (v * \phi)} = \ang{u, \tau_x \Rr(v * \phi)} = (u * (v * \phi))(x). 
	\end{align*}
\item Uniqueness: we have shown in the lectures that if $w * \phi = w' * \phi$ for all $\phi$, then $w = w'$. This shows that \cref{def:convolution} uniquely defines $u * v$. 
\end{enumerate}
%	\[
%	\ang{u * v, \phi} = ((u * v) * \refl\phi)(0) = (u * (v * \refl\phi))(0),
%	\]
%	which shows that the formula \cref{def:convolution} uniquely defines $\ang{u * v, \phi}$ for any $\phi \in \DD(\RR^n)$, and therefore that $u * v$ is well-defined. For uniqueness,

@@

	(??)

	Now, for $u \in H^s(\RR^n)$, let $(\phi_m) \to u$ in $L^2$. Then we have
	\begin{align*}
		\norm{\phi_m - u}_{H^s}^2 = \int \ang{\lambda}^{2s} \abs{(\phi_m - u)(\lambda)}^2 \dd{\lambda}
	\end{align*}

(??)
\end{proof}

\begin{question}
	Prove that multiplication by a Schwarz function gives rise to a continuous map from $H^s(\RR^n)$ to itself, i.e., $\norm{\phi u}_{H^s} \les \norm{u}_{H^s}$ for $\phi \in \SS(\RR^n)$. You may assume Peetre's inequality: for $\lambda, \nu \in \RR^n$ and $s \in \RR$
	\[
	\qty(\frac{1 + \norm{\lambda}^2}{1 + \norm{\mu}^2})^s \leq 2^{\abs{s}} \qty(1 + \norm{\lambda - \mu}^2)^{\abs{s}}. 
	\]
\end{question}

\begin{proof}
	We have
	\begin{align*}
		\norm{\phi u}_{H^s}^2 &= \int \abs{\FF[\phi u](\lambda)}^2 (1 + \norm{\lambda})^2 \dd{\lambda}
	\end{align*}
(??) (how to bound fourier transform of product??)
\end{proof}
%-----------------------------------------------------------------------------
\setcounter{question}{18}

### 2020-11-18 modern_statistical_methods/C3.tex
aaarhhh
hellow

Furthermore, when the spectrum of $\Sigma$ is concentrated on a few directions, the rate can be improved significantly. For this, we need a definition.
\begin{definition}
	The \emph{effective rank} $r(\Sigma)$ of a nonzero symmetric positive semi-definite matrix $\Sigma$ with eigenvalues $0 \leq \gamma_d(\Sigma) \leq \dotsb \leq \gamma_1(\Sigma)$ is given by
	\[
	r(\Sigma) = \frac{\Tr(\Sigma)}{\norm{\Sigma}_\Rm{op}} = \frac{\sum_{i=1}^d \gamma_i(\Sigma)}{\gamma_1(\Sigma)}. 
	\]
\end{definition}
It is easily seen that $r(\Sigma) \leq \Rm{rank}(\Sigma)$, with equality if and only if all nonzero eigenvalues of $\Sigma$ are equal. The following theorem (which we will not prove) gives a tighter bound if we can assume that the rows of $X$ are normally distributed and that $r(\Sigma)$ is constrained:
\begin{theorem}
	Suppose $X$ has i.i.d.\ $N(0, \Sigma)$ rows with $r(\Sigma) \leq cn$ (??). Then there exists $C$ such that for all $\delta > 0$ we have
	\[
	\PP\qty(\frac{\norm{\hat\Sigma - \Sigma}_\Rm{op}}{\norm{\Sigma}_\Rm{op}} \leq C \qty(\frac{r(\Sigma) + \delta}{n} \lor \sqrt{\frac{r(\Sigma) + \delta}{n}})) \geq 1 - e^{-\delta}. 
	\]
\end{theorem}

@@

>>>>

Furthermore, when the spectrum of $\Sigma$ is concentrated on a few directions, the rate can be improved significantly. For this, we need a definition.
\begin{definition}
	The \emph{effective rank} $r(\Sigma)$ of a nonzero symmetric positive semi-definite matrix $\Sigma$ with eigenvalues $0 \leq \gamma_d(\Sigma) \leq \dotsb \leq \gamma_1(\Sigma)$ is given by
	\[
	r(\Sigma) = \frac{\Tr(\Sigma)}{\norm{\Sigma}_\Rm{op}} = \frac{\sum_{i=1}^d \gamma_i(\Sigma)}{\gamma_1(\Sigma)}. 
	\]
\end{definition}
It is easily seen that $r(\Sigma) \leq \Rm{rank}(\Sigma)$, with equality if and only if all nonzero eigenvalues of $\Sigma$ are equal. The following theorem (which we will not prove) gives a tighter bound if we can assume that the rows of $X$ are normally distributed and that $r(\Sigma)$ is constrained:
\begin{theorem}
	Suppose $X$ has i.i.d.\ $N(0, \Sigma)$ rows with $r(\Sigma) \leq cn$. Then there exists $C$ (depending on $c$) such that for all $\delta > 0$ we have
	\[
	\PP\qty(\frac{\norm{\hat\Sigma - \Sigma}_\Rm{op}}{\norm{\Sigma}_\Rm{op}} \leq C \qty(\frac{r(\Sigma) + \delta}{n} \lor \sqrt{\frac{r(\Sigma) + \delta}{n}})) \geq 1 - e^{-\delta}. 
	\]
\end{theorem}

@@

\subsubsection{The graphical Lasso}
We saw that if we make the structural assumption that the effective rank of $\Sigma$ is small, the sample covariance matrix $\hat\Sigma$ is a good estimator. However, there are other possible structural assumptions. 
One is based on the following:
\begin{proposition}
	Let $Z \sim N(\mu, \Omega^{-1})$. Then $z_k \perp z_j \mid z_{-jk}$ ($z_k$ is independent of $z_j$ conditional on the other $z_i$) if and only if $\Omega_{jk} =0 $.  
\end{proposition}

\begin{proof}
	We have
	\begin{align*}
		f(z_j, z_k \mid z_{-jk}) &= \frac{f(z_1, \dotsc, z_n)}{f(z_{-jk})} \propto f(z_1, \dotsc, z_n) \propto \exp(-\frac12 (z-\mu)\T \Omega(z-\mu)) \\
		&= g(z_j) h(z_k) \exp(- z_j z_k \Omega_{jk}), 
	\end{align*}
and this factorises if and only if $\Omega_{jk} = 0$. 
\end{proof}

Assuming that many variables are independent of each other conditional on the other variables (or in other words, that $z_k$ and $z_j$ depend on each other only ``through the other $z_i$''), is a fairly natural assumption, which means it is sensible to assume that $\Omega$ is sparse. If this is the case, $\hat\Sigma$ may not be the best estimator.

The \emph{graphical lasso} explicitly penalises estimators of $\Omega$ with many non-zero entries, and is defined as
\[
\hat\Omega^\Rm{GL} \ceq \argmin_{\Omega \succ 0} \qty(-\log(\det\Omega) + \Tr(\hat\Sigma\Omega) + \lambda \norm{\Omega}_1), 
\]
where the 1-norm of a matrix is defined as the sum of the absolute values of its entries. This is exactly minus the log-likelihood plus a penalty term, and finding $\hat\Omega^\Rm{GL}$ is a convex optimisation problem which generally leads to sparse estimators.

Note that when we penalise $\norm{\Omega}_1$, we are also penalising the diagonal entries, which have nothing to do with conditional independence. Therefore, sometimes we penalise only $\sum_{j \neq k} \abs{\Omega_{jk}}$.

\subsection{Principal Component Analysis (PCA)}
Let $X \in \RR^{n\times d}$ have i.i.d.\ rows with mean 0 and covariance $\Sigma$. Let $\lambda_1 \geq \dotsb \geq \lambda_d \geq 0$ be the eigenvalues of $\Sigma$ with eigenvectors $u_1, \dotsc, u_ d\in S^{d-1}$. For any $s \leq d$ we have
\[
\Span(u_1, \dotsc, u_s) = \argmax_{\dim(S) = s} \EE[ \norm{P_S x_1}^2] = \argmin_{\dim(S) = s} \EE[\norm{x_1 - P_S x_1}^2], 
\]
i.e., $\Span(u_1, \dotsc, u_s)$ is the $s$-dimensional subspace of $\RR^d$ with maximum variance and minimal residual variance.

The quantity $q_s = \frac{\sum_{i=1}^s \lambda_i}{\Tr(\Sigma)}$ is the proportion of variance captured by the largest $s$ eigenvalues, and this may be close to 1 for $s \ll d$. The idea of PCA is to approximate the eigenvalues and eigenvectors of $\Sigma$ by those of $\hat\Sigma = \frac1n (X - \bar X \mathbf{1})\T (X - \bar X \mathbf{1}) \ceq \frac1n \tilde X\T \tilde X$.

The main question is: how robust are eigenvectors and eigenvalues under perturbation of the data?

\subsubsection{Perturbation of eigenvalues and eigenvectors}
Let $E = \hat\Sigma - \Sigma$, and suppose $E$ is a deterministic perturbation. We have proven bounds on $\norm{E}_\Rm{op}$. It is easily seen that
\[
\hat{\lambda_1} = \max_{v \in S^{d-1}} v\T (\Sigma + E)v \leq\max_{v \in S^{d-1}} v\T \Sigma v + \max_{v \in S^{d-1}} v\T E v  \leq \lambda_1 + \norm{E}_\Rm{op}, 
\]
and analogously it can be seen that $\lambda_1 \leq \hat\lambda_1 + \norm{E}_\Rm{op}$, so we obtain $\abs{\lambda_1 - \hat\lambda_1} \leq \norm{E}_\Rm{op}$. This can be generalised:
\begin{theorem}[Weyl]
	We have $\max_i \abs*{\hat\lambda_i - \lambda_i} \leq \norm{E}_\Rm{op}$. 
\end{theorem}

We conclude that eigenvalues are generally robust to small perturbations. However, eigenvectors are genearlly not robust to small perturbations. This is because when $\lambda_i$ and $\lambda_{i+1}$ are close to each other, then $\Span(\hat u_i, \hat u_{i+1})$ may be a good approximation of $\Span(u_i, u_{i+1})$, but it is not possible to identify the individual eigenvectors.  Therefore we must place further assumptions on the eigenvectors.

### 2020-11-18 topics_statistical_theory/C3.tex
hellow

Suppose $x \geq 1 - h\Delta$, then we have $u_1 \leq -\Delta$ and $u_n \geq 0$, and we can apply very similar arguments to reach the desired conclusion. 
\end{proof}
%Suppose that $X\T W X$ is positive definite (this is a very weak condition), then we have
%\begin{align*}
%	\hat\beta = \argmin_{\beta \in \RR^{p+1}} (\beta - (X\T W X)^{-1} X\T W Y)\T (X\T W X) (\beta - (X\T W X)^{-1} X\T W Y),
%\end{align*}
%which is minimised at $\hat\beta = (X\T W X)^{-1} X\T WY$.

>>>>

Suppose $x \geq 1 - h\Delta$, then we have $u_1 \leq -\Delta$ and $u_n \geq 0$, and we can apply very similar arguments to reach the desired conclusion. 
\end{proof}

Now that we have bounds on the variance and bias, we can prove our uniform bound: 
\begin{theorem}
	Under the conditions of the previous two propositions, if we choose $h = \alpha n^{-1/(2\beta + 1)}$ for some $\alpha > 0$, there exists $n_0 \in \NN, C > 0$, depending only on $\beta, L, \alpha, K, \sigma_\Rm{max}^2$, such that
	\[
	\sup_{m \in \HH(\beta, L)} \sup_{x_0 \in [0, 1]} \EE\qty[ \qty{\hat m_n(x_0;p) - m(x_0)}^2] \leq Cn^{-2\beta/(2\beta + 1)}.
	\]
\end{theorem}

%Suppose that $X\T W X$ is positive definite (this is a very weak condition), then we have
%\begin{align*}
%	\hat\beta = \argmin_{\beta \in \RR^{p+1}} (\beta - (X\T W X)^{-1} X\T W Y)\T (X\T W X) (\beta - (X\T W X)^{-1} X\T W Y),
%\end{align*}
%which is minimised at $\hat\beta = (X\T W X)^{-1} X\T WY$.

\subsection{Splines}
\subsubsection{Cubic splines}
Let $n \geq 3$ and $a \leq x_1 < \dotsb < x_n \leq b$.  
\begin{definition}
	A function $g \colon [a, b] \to \RR$ is called a \emph{cubic spline} with \emph{knots} $x_1, \dotsc, x_n$ if 
	\begin{enumerate}
		\item $g$ is a cubic polynomial on each interval $(a, x_1), (x_1, x_2), \dotsc, (x_n, b)$;
		\item $g \in C^2[a, b]$. 
	\end{enumerate}
	Furthermore, $g$ is called \emph{natural} if it is linear on $[a, x_1]$ and $[x_n, b]$, (i.e., $g''(a) = g'''(a) = g''(b) = g'''(b) = 0$). 
\end{definition}

We often represent a natural cubic spline by the vectors $\vec g \in \RR^n$ with $g_i = g(x_i)$, and $\vec \gamma \in \RR^{n-2}$ with $\gamma_i = g''(x_i)$ (excluding $\gamma_1$ and $\gamma_n$). Writing $h_i \ceq x_{i+1} - x_i$ we have for $x \in [x_i, x_{i+1}]$ that
\[
g(x) = \frac{(x - x_i) g_{i+1} - (x_{i+1} - x) g_i}{h_i} - \frac16(x - x_i)(x_{i+1} - x) \qty{\qty(1 + \frac{x - x_i}{h_i})\gamma_{i+1} + \qty(1 + \frac{x_{i+1} - x}{h_i}\gamma_i)}.
\]
\begin{proposition}
	Given $\vec g \in \RR^n$, there exists a unique natural cubic spline $g$ with knots at $x_1, \dotsc, x_n$ satisfying $g(x_i) = g_i$ for all $i$, and there exists $K \succeq 0$ (depending on $x_1, \dotsc, x_n$) such that
	\[
	\int_a^b g''(x)^2 \dd{x} = \vec g\T K \vec g. 
	\]
\end{proposition}
We call $g$ the \emph{natural cubic spline interpolant to $\vec g$ at $x_1, \dotsc, x_n$}.

\begin{definition}
	We define $\SS_2[a, b]$ as the set of real-valued functions on $[a, b]$ with an absolutely continuous first derivative. For $f \in \SS_2[a, b]$, we define the \emph{roughness} of $f$ by $R(f) \ceq \int_a^b f''(x)^2 \dd{x}$. 
\end{definition}

\begin{proposition}
	For any $\vec g \in \RR^n$, the natural cubic spline interpolant to $\vec g$ at $x_1, \dotsc, x_n$ is the unique minimiser of $R$ over all $g \in \SS_2[a, b]$ that satisfy $g(x_i) = g_i$ for all $i$. 
\end{proposition}

\subsubsection{Natural cubic smoothing splines}
Consider the nonparametric regression model $Y_i = g(x_i) + \sigma \eps_i$, where $\EE[\eps_i] = 0$ and $\Var[\eps_i] = 1$. A way to estimate a nonparametric regression function is to balance data fidelity against roughness of the curve, which can be done by minimising 
\[
S_\lambda(g) \ceq \sum_{i=1}^n (Y_i - g(x_i))^2 + \lambda R(g),
\]
where $\lambda > 0$ is a regularisation parameter. For small $\lambda$, this is almost an exact fit to the data. For large $\lambda$, we are basically minimising $\abs{g''}$, which means we will approximate the linear regression fit.

\begin{theorem}
For each $\lambda \in (0, \infty)$, there exists a unique minimiser $\hat g_\lambda$ of $S_\lambda$ over $\SS_2[a, b]$. It is the natural cubic spline with knots at $x_1, \dotsc, x_n$ and $\vec g = (I + \lambda K)^{-1} \vec Y$. 
\end{theorem}

\begin{proof}
	If $g$ is not a natural cubic spline, we know that the natural cubic spline $g^*$ which interpolates $g(x_1), \dotsc, g(x_n)$ at $x_1, \dotsc, x_n$ has a strictly lower value of $S_\lambda$, so we know the minimiser must be a natural cubic spline.

	If $g$ is a natural cubic spline, then there exists $K \succeq 0$ such that
	\begin{align*}
		S_\lambda(g) &= (\vec Y - \vec g)\T (\vec Y - \vec g) + \lambda \vec g\T K \vec g \\
		&= \vec g\T (I + \lambda K)\vec g - 2 \vec Y\T \vec g + \vec Y\T\vec Y,
	\end{align*}
and by ``completing the square'' we write, for some $Z$ independent of $\vec g$, 
\[
S_\lambda(g) = \qty(\vec g - (I + \lambda K)^{-1} \vec Y)\T (I + \lambda K) \qty(\vec g - (I + \lambda K)^{-1} \vec Y) + Z
\]
Since $I + \lambda K$ is positive definite it follows that $\vec g = (I + \lambda K)^{-1} \vec Y$ gives the minimiser. 
\end{proof}
